A tool set for random number generation on GPUs in R

### 2021-11-27 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3005 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Now we use the streams created in section \ref{createstreams} to generate a vector of double-precision i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
as.vector(clrng::runif(n=6, streams=myStreamsGpu, Nglobal=c(2,2), type = "double"))
## [1] 0.7353245 0.8423426 0.6142074 0.2159195 0.1100781 0.8698300
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}
%\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu} and \code{myStreamsGpu2} here. 
Reusing \code{myStreamsGpu} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu} after it being used.

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.073   0.000   0.072
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.658   1.125   7.779
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection{Exponential random number generation}
The Exponential random variates are produced by applying the inverse transform method on i.i.d. $U(0,1)$ random numbers. The random variable $X \sim \text{Exponential}(\lambda)$ has cumulative distribution function $F_{X}(x)=1-e^{-\lambda x}$ for $x \geq 0$ and $\lambda > 0$. The inverse of $F_{X}$ is $F^{-1}_{x}(y)= -(1/ \lambda) \log (1-y)$, for $ 0 \leq y <1$. The kernel is not shown because it is mostly same as the kernel for uniform and normal random numbers, except for the part that does the inverse transform. %Developers can make use of our kernel for generating other types of random variables on GPU in \proglang{R}.

Below is an example that we produce a matrix of exponential random number with expectation 1.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
r_matrix <- clrng::rexp(c(2,4), rate=1, Nglobal=c(2,2), type="double")
as.matrix(r_matrix)
##            [,1]      [,2]     [,3]       [,4]
## [1,] 0.05469212 0.6969336 1.089906 0.78689966
## [2,] 0.04188041 0.1686520 1.905295 0.03536933
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

The arguments of the \fct{clrng::fisher.sim} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{x}: a ``vclMatrix''.
  \item \code{N}: number of simulation runs.
  \item \code{streams}: streams.
  \item \code{type}: ``double'' or ``float'' of returned test statistics.
  \item \code{returnStatistics}: if TRUE, return all test statistics. 
  \item \code{Nglobal}: global index space. Default is set as $(64,16)$.
\end{itemize}
Users request $N$ number of replicates, while the actual number of replicates to be executed on GPU is \code{ceiling(N/prod(Nglobal))*prod(Nglobal)}. We show the advantage of \fct{clrng::fisher.sim} by computing the p-values for two real data examples: one with a relatively big p-value and another with a very small p-value, and compare the run-time with using \fct{stats::fisher.test} for each of the data sets on two computers: one with a very good CPU and an ordinary GPU, the other is equipped with an excellent GPU and an ordinary CPU. The \proglang{R} outputs for testing on computer 2 is shown in the following.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n =256*64, initial=666)
# almost.1 <- 1 + 64 * .Machine$double.eps
# observed_m= -logfactSum(month, c(16,16))/almost.1

month_gpu<-vclMatrix(month,type="integer")
system.time(result_month <- clrng::fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,64)))
##    user  system elapsed 
##   0.308   0.008   0.316
result_month$threshold
## [1] -47954.8
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410825
result_month$p.value
## [1] 0.4044323
\end{verbatim}
\end{kframe}
\end{knitrout}
We got 409524 cases whose test statistics are below the observed threshold based on an actual number of 1015808 simulations, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}. \fct{clrng::fisher.sim} takes about 0.69 seconds, \pkg{clrng} accounts for 4.64\% of the elapsed time on CPU.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#observed_w= -clrng::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_week<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
##    user  system elapsed 
##   1.937   0.011   1.947
result_week$threshold
## [1] -54989.56
result_week$simNum
## [1] 10010624
result_week$counts
## [1] 1193
result_week$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

@@

\includegraphics[width=0.45\textwidth]{figure/fighistMonth-1} \includegraphics[width=0.45\textwidth]{figure/fighistMonth-2} \end{kframe}
\end{knitrout}
\caption{Approximate sampling distributions of the test statistics from the two examples ``Month'' and ``Week''. The test statistic values of the observed tables are calculated by \fct{clrng::logfactSum} and are marked with a blue line on each plot. \label{fig4}}
\end{figure}

>>>>

Now we use the streams created in section \ref{createstreams} to generate a vector of double-precision i.i.d. $U(0,1)$ random numbers with \fct{clrng::runif}. To view the generated random numbers, we need to convert them to \proglang{R} vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
as.vector(clrng::runif(n=6, streams=myStreamsGpu, Nglobal=c(2,2), type = "double"))
## [1] 0.7353245 0.8423426 0.6142074 0.2159195 0.1100781 0.8698300
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}
%\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu} and \code{myStreamsGpu2} here. 
Reusing \code{myStreamsGpu} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu} after it being used.

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.071   0.003   0.073
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.658   1.125   7.779
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection{Exponential random number generation}
The exponential random variates are produced by applying the inverse transform method on i.i.d. $U(0,1)$ random numbers. The random variable $X \sim \text{Exponential}(\lambda)$ has cumulative distribution function $F_{X}(x)=1-e^{-\lambda x}$ for $x \geq 0$ and $\lambda > 0$. The inverse of $F_{X}(\cdot)$ is $F^{-1}_{X}(y)= -(1/ \lambda) \log (1-y)$, for $ 0 \leq y <1$. The kernel is not shown because it is mostly same as the kernel for uniform and normal random numbers, except for the part that does the inverse transform. %Developers can make use of our kernel for generating other types of random variables on GPU in \proglang{R}.

Below is an example that produces a matrix of Exponential random numbers with expectation equal to 1.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
r_matrix <- clrng::rexp(c(2,4), rate=1, Nglobal=c(2,2), type="double")
as.matrix(r_matrix)
##            [,1]      [,2]     [,3]       [,4]
## [1,] 0.05469212 0.6969336 1.089906 0.78689966
## [2,] 0.04188041 0.1686520 1.905295 0.03536933
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

The arguments of the \fct{clrng::fisher.sim} are described as follows:
\begin{itemize}
\setlength{\parskip}{0pt}
  \item \code{x}: a contingency table, a ``vclMatrix'' object.
  \item \code{N}: number of simulation runs.
  \item \code{streams}: streams.
  \item \code{type}: ``double'' or ``float'' of returned test statistics.
  \item \code{returnStatistics}: if TRUE, return all test statistics. 
  \item \code{Nglobal}: global index space. Default is set as $(64,16)$.
\end{itemize}
Users request $N$ number of replicates, while the actual number of replicates to be executed on GPU is larger (\code{ceiling(N/prod(Nglobal))*prod(Nglobal)}). We show the advantage of \fct{clrng::fisher.sim} by computing the p-values for two real data examples: one with a relatively big p-value and another with a very small p-value, and compare the run-time with using \fct{stats::fisher.test} for each of the data sets on two computers: one with a very good CPU and an ordinary GPU, the other is equipped with an excellent GPU and an ordinary CPU. The \proglang{R} outputs for testing on computer 2 is shown in the following.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n =256*64, initial=666)
month_gpu<-vclMatrix(month,type="integer")
system.time(result_month <- clrng::fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,64)))
##    user  system elapsed 
##   0.315   0.000   0.315
result_month$threshold
## [1] -47954.8
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410825
result_month$p.value
## [1] 0.4044323
\end{verbatim}
\end{kframe}
\end{knitrout}
We got 410825 cases whose test statistics are below the observed threshold based on an actual number of 1015808 simulations, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}. \fct{clrng::fisher.sim} takes about 0.31 seconds, \pkg{clrng} accounts for 2.17\% of the elapsed time on CPU.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_week<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
##    user  system elapsed 
##   1.968   0.022   1.991
result_week$threshold
## [1] -54989.56
result_week$simNum
## [1] 10010624
result_week$counts
## [1] 1193
result_week$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

@@

\includegraphics[width=0.45\textwidth]{figure/fighistMonth-1} \includegraphics[width=0.45\textwidth]{figure/fighistMonth-2} \end{kframe}
\end{knitrout}
\caption{Approximate sampling distributions of the test statistics from the two examples ``Month'' and ``Week''. The test statistic values of the observed tables are marked with a blue line on each plot. \label{fig4}}
\end{figure}
